---
title: "Cassie's objectives"
author: "Cassie Pagnozzi"
output: pdf_document
date: "2024-09-19"
---
###Formatting RMD files: 
To start any R Markdown file, you must always begin with a header that includes basic information such as the title, author, output, and date. This header gives context about who is providing the file and what it's about. A typical header is shown above. 
Depending on what format you need, you can replace the output with html_document, word_document, or whatever document you need for the format.  

###Running libraries:
Before running any code in R, it is recommended that you load necessary packages that you may need before starting your analyses. 
```{r Load Libraries, include=FALSE}
# Load other packages here.
if (!require("cowplot")) install.packages("cowplot"); library(cowplot)
if (!require("conflicted")) install.packages("conflicted"); library(conflicted) # For dealing with conflicts
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse)# For everything
if (!require("viridis")) install.packages("viridis"); library(viridis)
conflict_prefer_all("dplyr", quiet = TRUE)
```

###Using code chunks
In an RMD file, code chunks are blocks of code that read commands. This is indicated by the green button with a plus sign in the upper right corner of the console. Code chunks will always begin with a title and three back ticks and end with back ticks. The code chunk will not run properly if this is not indicated correctly.  
To run a code chunk, hit the third symbol, signified by a green arrow in the top right corner of the single chunk. 
To run every previous code chunk, hit second button in the top right corner of the code chunk, this is indicated by a downwards arrow. 
```{r Fun fact}
#You can do math using code chunks
2+8
```
#As I have done previously, you can make comments all throughout an RMD file by beginning your statements with a hashtag (#) This prevents R from reading it as code, and instead as text. 
```{r}
What happens when you run code with no hashtags? Try running this code chunk to see what happens!
penguins <- read.csv("palmerpenguins_extended.csv")
```

Error! 
```{r Reading in our data}
#Lets try that again:
penguins <- read.csv("palmerpenguins_extended.csv")
```
Success!
After hitting the green arrow for the particular code chunk, you can always check in the console below if you have properly read in the file. 

###Assessing normality and determining what statistical test to use
Depending on what data you are working with and what you want to find, you can use various kinds of statistical tests within your RMD file. 
  T-Tests are great for comparing the mean of the data. There are multiple kinds of t-tests you can conduct, depending on what you want to compare the mean with. 
```{r Two Sample T-Test}
#For this example, we will be comparing the average body mass of male and female penguins, therefore our variables will be body mass and sex. 
t.test(body_mass_g ~ sex, data = penguins)
```

  Another statistical test are ANOVA, which measures the variance for multiple groups. 
```{r}
#Here, we can test for the effect of species and sex on body mass in pengiuns
aov(body_mass_g ~ species * sex, data = penguins)
```

  Linear regression models are also great statistical tests for estimating the linear relationship between one or more explanatory variables. 
```{r Linear Regression Model}
multiple_lm <- lm(body_mass_g ~ flipper_length_mm + bill_length_mm, data = penguins)
summary(multiple_lm)
```
The output from this lm displays the estimate coefficients for flipper and bill length, as well as their p-values. This information can be really helpful for how different variables can determine body mass. 


#git config pull.rebase false - can use dropdown on pull menu to pull with 
#rebase automatically
#We can make branches off the project to prevent conflicts until the end 
## Guidelines:
#pull before starting to work/edit