---
title: "Group 4 Project 1"
authors: "Thomas Mazzarulli, Cassie Pagnozzi, and Hedy Davoudi Kianersi"
output: github_document
date: "2024-10-08"
---

## Target Audience
Our target audience are students in Biology 204. Some experiences with R but not advanced level. No experiences with markdown 

## Instructions and Accessing R Tutorial
Have RStudio downloaded and make sure it is updated to the latest version
Open GitHub on your web browser and access the Bucknell-Biol 364 Repository 
Search for Project_1_Group_4_2024, click the green code button on the top right of the repo, and click “download zip” to load the documents into your folder- move these files into a folder you can easily access
Open R Studio again and find these documents using the files tab in the bottom right corner of R
Click on Project01.Rmd to load the tutorial!


## Objectives
Learn how to properly use code chunks, make comments, and properly format an RMD file - Cassie
Generating data frames from imported data, understanding different variable and data types in R - Thomas 
Organizing and mutating/manipulating dataframes, removing outliers and N/A values Hedy
Assessing normality and determining which statistical tests to use- Cassie
Generating visually appealing graphs Hedy 
Combining figures with their respective statistical tests - Thomas

## Hypothesis/examples
The diet of penguins influences their mass
Penguins of the same species will have different mean body masses based on their location
The age of the penguin will affect it’s primary diet - will need to filter out chicks as they have a parental diet 
The island the penguins are located on affects the health condition of the penguins 

dHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHb
HHP%%#%%%%%%%%%%%%%%%%#%%%%%%%#%%VHH
HH%%%%%%%%%%#%v~~~~~~\%%%#%%%%%%%%HH
HH%%%%%#%%%%v'        ~~~~\%%%%%#%HH
HH%%#%%%%%%v'dHHb      a%%%#%%%%%%HH
HH%%%%%#%%v'dHHHA     :%%%%%%#%%%%HH
HH%%%#%%%v' VHHHHaadHHb:%#%%%%%%%%HH
HH%%%%%#v'   `VHHHHHHHHb:%%%%%#%%%HH
HH%#%%%v'      `VHHHHHHH:%%%#%%#%%HH
HH%%%%%'        dHHHHHHH:%%#%%%%%%HH
HH%%#%%        dHHHHHHHH:%%%%%%#%%HH
HH%%%%%       dHHHHHHHHH:%%#%%%%%%HH
HH#%%%%       VHHHHHHHHH:%%%%%#%%%HH
HH%%%%#   b    HHHHHHHHV:%%%#%%%%#HH
HH%%%%%   Hb   HHHHHHHV'%%%%%%%%%%HH
HH%%#%%   HH  dHHHHHHV'%%%#%%%%%%%HH
HH%#%%%   VHbdHHHHHHV'#%%%%%%%%#%%HH
HHb%%#%    VHHHHHHHV'%%%%%#%%#%%%%HH
HHHHHHHb    VHHHHHHH:%odHHHHHHbo%dHH
HHHHHHHHboodboooooodHHHHHHHHHHHHHHHH
HHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHH
VHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHV

```{r}
# Load other packages here.
if (!require("cowplot")) install.packages("cowplot"); library(cowplot)
if (!require("conflicted")) install.packages("conflicted"); library(conflicted) # For dealing with conflicts
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse) # For everything
if (!require("rstatix")) install.packages("rstatix"); library(rstatix) 

conflict_prefer_all("dplyr", quiet = TRUE)
```


## Penguin analysis R tutorial

Throughout this tutorial, we will introduce some of the basic properties and functions of R, as well as some additional packages which can be used to enhance it's functionality. In order to show these tools, we will be using the classic Palmer Penguins dataset, and exploring the relationships between many of the variables it includes.  


## Objective 1: Using code chunks

In an RMD file, code chunks are blocks of code that read commands. This is indicated by the green button with a plus sign in the upper right corner of the console. Code chunks will always begin with a title and three back ticks and end with back ticks. The code chunk will not run properly if this is not indicated correctly.  
To run a code chunk, hit the third symbol, signified by a green arrow in the top right corner of the single chunk. 
To run every previous code chunk, hit second button in the top right corner of the code chunk, this is indicated by a downwards arrow. 
```{r Fun fact}
#You can do math using code chunks
2+8
```

As I have done previously, you can make comments all throughout an RMD file by beginning your statements with a hashtag (#) This prevents R from reading it as code, and instead as text. 

```{r Reading in our data}
#Make sure you have the dataset file downloaded and can be easily accessed in your working directory before trying to run the code!
penguins <- read.csv("palmerpenguins_extended.csv")
```

Success!
After hitting the green arrow for the particular code chunk, you can always check in the console below if you have properly read in the file. If not, make sure you downloaded the dataset and that it is in your working directory.


## Objective 2: Understanding our data

Now that we have imported our data into R, it's important to check that everything is 
what we expect it to be. The CSV file that we imported has now taken the form of a dataframe in our R environment, similar to an Excel file. However, in R, each of the columns will be treated as a variable. Based on the characteristics of the data in each column, R will treat it as a certain type of variable. Using the str(penguins) function, we can get a brief summary of how R is interpreting this data.   

```{r}
str(penguins)
```

Based on the output of this command, we can see that R is interpreting variables like bill length, bill depth, flipper length, and body mass as numeric values. However, some values, such as species, sex, and island are being treated as characters. This means that R is treating them as strings of text, rather than as categories. Since our analysis requires these columns to be used as groups rather than individual strings, we need to tell R that these are categorical values, or factors. In this case, the variables species, island, sec, diet, life stage, and health metrics are all categories, rather than unique observations for each individual. 

```{r}
penguins$species <- as.factor(penguins$species)
penguins$island <- as.factor(penguins$island)
penguins$sex <- as.factor(penguins$sex)
penguins$diet <- as.factor(penguins$diet)
penguins$life_stage <- as.factor(penguins$life_stage)
penguins$health_metrics <- as.factor(penguins$health_metrics)
```

Now that we've told R how to treat these variables, we should make sure that everything is as we expect. We can use the str() function once again to see if R is interpreting our dataframe as intended. 

```{r}
str(penguins)
```

Now we can see that our data is being interpreted correctly by R. Before we run any analysis of this data, it's important to check that the dataset we're using is complete. If there are any N/A values present, they might cause an issue. The is.na() command will tell us if any missing values are present. 

```{r, results='hide'}
is.na(penguins)
```
Wow! This command left us with a lot to interpret, as it was simply checking each value in the dataframe. With more than 3000 observations, this isn't appropriate for our dataset. However, we can use the summary() command to give us a glimpse at each of our columns, which will let us know if anything is missing. 

```{r}
summary(penguins)
```

Based on this summary, it seems that we luckily aren't missing any observations. We can now move on to our analysis. 


## Objective 3: Manipulating our data

During data exploration, particularly with complex datasets, it is often necessary to manipulate the DataFrame. This may involve removing columns that do not provide significant information or grouping specific data of interest to facilitate a more efficient data analysis process. In order to do this we will show some functions which are part of tidyverse package.

The group_by function is utilized to examine the relationships among specific columns in a dataset. In this case, we aim to investigate the relationship between island location, dietary sources, and body mass. Our hypothesis posits that the geographical location of an island may influence the dietary habits of the animals, subsequently impacting their body mass.
However, before diving into the group function it is important to talk about one one the singnificant R functions which is pipeline operator |>
The pipe operator takes the output of one expression and passes it as the first argument to the next expression. This makes it easier to read and understand the flow of data transformations.
•  Readability: The code reads more like a sequence of operations, making it easier to follow.
•  Less Nesting: It reduces the need for nested function calls, which can be hard to read.
•  Debugging: You can easily inspect the output at each step if needed, making debugging simpler.


```{r groupby}
penguins |> 
  group_by(island, diet, body_mass_g) |>
  summarize(meanMass=mean(body_mass_g, na.rm = TRUE))
```

By running the code chunk above we can see that interpretation of the data became much easier.  

Next, we will incorporate an additional column into our dataset. For example, to assess the Body Condition Index (BCI), we should divide the total body mass by the lengths of the body parts. To do so we need to use mutate function which again is one part of tidyverse package.

```{r mutate}
penguins <- mutate(penguins, BCI = body_mass_g / flipper_length_mm)
summary(penguins$BCI)
```

By running the code chunk above we can see that we added an extra column to our existed data set named BCI.

Next, when want to make subsets of rows from a data frame we can use filter function. This allows us to make another dataset which make the data interpretation easier. 
Here by running the code chunk we can see that we made another data set called palmerpenguins_extended.filter which has 2826 observations with 12 variables and we can see there 2826 observations for penguins under 6038 g.

```{r filter}
penguins.filter <- filter(penguins, body_mass_g < 6038)
dim(penguins.filter)
```

dim() function is used to retrieve or set the dimensions of an object, such as a matrix or an array. It returns a vector containing the size of each dimension of the object.
•	Works with matrices, arrays, and data frames.
•	For data frames, dim() will return the number of rows and columns.
•	It’s useful for understanding the structure of your data in R.


## Objective 4: Assessing normality and determining what statistical test to use

Depending on what data you are working with and what you want to find, you can use various kinds of statistical tests within your RMD file. 
  T-Tests are great for comparing the mean of the data. There are multiple kinds of t-tests you can conduct, depending on what you want to compare the mean with.
  
```{r Two Sample T-Test}
#For this example, we will be comparing the average body mass of male and female penguins, therefore our variables will be body mass and sex. 
t.test(body_mass_g ~ sex, data = penguins)
```

These tests can output useful information, such as a p-value, which can tell you if there is a significant difference. 

  Another statistical test are ANOVA, which measures the variance for multiple groups. 
  
```{r}
#Here, we can test for the effect of species and sex on body mass in penguins
aov(body_mass_g ~ species * sex, data = penguins)
```

  Linear regression models are also great statistical tests for estimating the linear relationship between one or more explanatory variables. 
  
```{r Linear Regression Model}
multiple_lm <- lm(body_mass_g ~ flipper_length_mm + bill_length_mm, data = penguins)
summary(multiple_lm)
```

The output from this lm displays the estimate coefficients for flipper and bill length, as well as their p-values. This information can be really helpful for how different variables can determine body mass. 

## Objective 5: Graphing

Analytical graphics play a vital role in data analysis, making it essential to select the appropriate method for visualizing our data.
making histograms with the data that you want to explore is one of the easiest way to visualize your data and to see if the data has the normal distribution. 
By looking at our code chunk we can see that we used the color, fill, and bins commands to make our histogram better and easier to interpret the data.

```{r histogram}
ggplot(penguins) +
  aes(x=body_mass_g) +
  geom_histogram(color = "black", fill = "white", bins = 30)
```

ggplot is a powerful tool in R to make more sophisticated graphs. Here we will break down this function to details to make it more clear.

Fist we should start by choosing the data frame and the variable that we want to visualize 

Aesthetics (aes): Defines how data variables map to visual properties (e.g., x and y axes, colors, shapes). For example: aes(x = variable1, y = variable2)

geom_ functions define the type of plot (e.g., points, lines, bars) and common geoms include:
geom_point(): scatter plots
geom_line(): line graphs
geom_bar(): bar charts
geom_histogram(): histograms (which we saw in the code chunk above)

Facets function splits data into subplots based on a factor which includes facet_wrap(~ variable) or facet_grid(rows ~ columns).

Finally, to make our plots more presentable we can use different themes to customize the appearance of the plot such as: Theme_ cowplot(), theme_minimal(), theme_classic(), 

Here lets see these functions step by step.

```{r}
ggplot(penguins) +
  aes(x = health_metrics) +
  geom_bar() +
  theme_cowplot()
```

Running the code chunk above allows us to compare the number of individuals based on their health metrics.

Now lets make our plot more professional.
If we wand to see the relationship between diet and body mass we can make a box plot as it is shown below.
we can use fill function to add color to our plot and differentiate among our variables also we can chose to remove the outliers by using this command (outlier.shape = NA)
We can see that our graph is much more presentable.

```{r}
ggplot(penguins) +
  aes(x = diet, y = body_mass_g, fill = diet) +
  geom_boxplot(outlier.shape = NA) +
  theme_cowplot()
```

Also, we can add another variable to the graph. For example, here I wanted to see how the health metrics looks like in each dietary categories and how they can contribute to body mass.
So, for the fill function I chose health_metrics

```{r}
ggplot(penguins) +
  aes(x = diet, y = body_mass_g, fill = health_metrics) +
  geom_boxplot(outlier.shape = NA) +
  theme_cowplot()
```

Finally, we can also make a line graph which allows us to visualize and interpret the relationship among three variables: body mass, BCI, and diet.
Here, we can see that labs function allows us to change the label of the graph and we can name each variable as we want them to be.

```{r}
ggplot(penguins) +
  aes(x = body_mass_g, y = BCI) +
  geom_point(aes(color = factor(diet))) +
  geom_smooth(method=lm , color="black", se=TRUE) +
  theme_cowplot() +
  labs(color="Diet") +
  xlab("body mass (g)") + 
  ylab("BCI")
```


## Objective 6: Combining stats and graphs

Now that we have some statistical analysis of our data, as well as some figures to better visualize it, the next step is to combine the two to allow for an easy interpretation. Using the package ggpubr, we can combine our statistical tests with our graphs to create some publication-quality figures. 

In order to accomplish this, we can use the package ggpubr. This package allows for us to complete both the functions of ggplot and our statistical analysis within the same section of code. The syntax in this package starts out the same as ggplot as shown previously, but we can add the function "stat_compare_means" to add the results of a statistical test and it's significance on to the chart. In the argument of this function, we have to specify the comparisons we want to provide the statistical tests for using the "comparisons" argument. We then have to specify which statistical method we want to use, using the "methods" argument, as well as how we want the results to be displayed, using the argument "label". Here is an an example showing the graph of the mass of penguins versus their diet categories.  

```{r}
if (!require("ggpubr")) install.packages("ggpubr"); library(ggpubr)

ggplot(penguins) +
  aes(x = diet, y = body_mass_g, fill = diet) +
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(comparisons = list(
      c("fish", "krill"), 
      c("fish", "parental"), 
      c("fish", "squid"),
      c("krill", "parental"), 
      c("krill", "squid"), 
      c("parental", "squid")),
      method = "t.test",
      label = "p.format") +  
  theme_cowplot()

```

This figure displays the p-values between each of our dietary groups, and it seems like there's a significant difference in weight for each of them! We can tidy this figure up a bit more by changing the labels. Additionally, using the viridis package, we can make the colors used in this figure more appropriate in cases of colorblindness, or if it needs to be printed in black and white. 

```{r}
if (!require("viridis")) install.packages("viridis"); library(viridis)

ggplot(penguins) +
  aes(x = diet, y = body_mass_g, fill = diet) +
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(comparisons = list(
      c("fish", "krill"), 
      c("fish", "parental"), 
      c("fish", "squid"),
      c("krill", "parental"), 
      c("krill", "squid"), 
      c("parental", "squid")),
      method = "t.test",
      label = "p.format") +  
  theme_cowplot() + 
  xlab("Diet") + 
  ylab("Body Mass (g)") + 
  scale_fill_viridis_d("diet")
```
This figure looks good, but the addition of a legend takes up a lot of space, and doesn't really provide any additional information. As a final modification, we can remove it and generate our final figure. 

```{r}

ggplot(penguins) +
  aes(x = diet, y = body_mass_g, fill = diet) +
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(comparisons = list(
      c("fish", "krill"), 
      c("fish", "parental"), 
      c("fish", "squid"),
      c("krill", "parental"), 
      c("krill", "squid"), 
      c("parental", "squid")),
      method = "t.test",
      label = "p.format") +  
  theme_cowplot() + 
  xlab("Diet") + 
  ylab("Body Mass (g)") + 
  scale_fill_viridis_d("diet") +
theme(legend.position = "none")  
```

Now, we have a visually appealing and easy to read figure that summarizes the relationship between the body mass of the penguins sampled and their respective diets, as well as the statistical significance of the differences between each of these categories. However, displaying these numbers in scientific notation gives us quite a bit of chart junk. Luckily, this package allows us to simplify, by specifying that labels should only be added when p values are significant, changing the labels argument from "label = 'p.format'" to "label = 'p.signif'". This allows us to display our p values using the standard notation of asterisks, rather than displaying the entire value on our chart. 

```{r}
ggplot(penguins) +
  aes(x = diet, y = body_mass_g, fill = diet) +
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(comparisons = list(
      c("fish", "krill"), 
      c("fish", "parental"), 
      c("fish", "squid"),
      c("krill", "parental"), 
      c("krill", "squid"), 
      c("parental", "squid")),
      method = "t.test",
      label = "p.signif") +  
  theme_cowplot() + 
  xlab("Diet") + 
  ylab("Body Mass (g)") + 
  scale_fill_viridis_d("diet") +
theme(legend.position = "none")  
```

This example should outline the syntax and format used to create high-quality figures. Using the examples we've provided so far in the tutorial, can you create a figure that compares groups and displays the statistical significance of the comparison? Some interesting comparisons to consider may be the island penguins are found on versus their mass, or investigating the difference between properties such as bill and flipper measurements between species. The following code should provide a skeleton, where you can replace the sections with three question marks (???) with the variables you are interested in examining. The formatting for this figure has been excluded, to allow you to play with different label and theme options. An example that shows this skeleton code filled in will follow the acknowledgements. 

```{r, eval = FALSE}

ggplot(penguins) +
  aes(x = ???, y = ???, fill = ???) +
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(comparisons = list(
      c("???", "???"), 
      c("???", "???"), 
      c("???", "???")), method = "???", label = "???") 
```

## Acknowledgements

For this project, we used the extended Palmer Penguins dataset.

Samy Baladram. (2023). 🐧 Palmer Penguins Dataset Extended [Data set]. Kaggle. https://doi.org/10.34740/KAGGLE/DS/3891364 

Additionally, we used information from the following textbooks.

Peng, R. D. (2015). Exploratory data analysis with r. Leanpub. http://leanpub.next/exdata

Touchon, J. C. (2021). Applied statistics with R: A practical guide for the life sciences (First edition.). Oxford University Press.

Throughout this tutortial, Hedy was the leader for sections 3 and 5, Cassie was in charge of 1 and 4, and Thomas worked on sections 2 and 6 and combined the rmd. 

## Example figure for final exercise

```{r}
ggplot(penguins) +
  aes(x = species, y = bill_length_mm, fill = species) +
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(comparisons = list(
      c("Adelie", "Gentoo"), 
      c("Adelie", "Chinstrap"), 
      c("Gentoo", "Chinstrap")), method = "t.test", label = "p.format") +
  theme_cowplot() + 
  xlab("Species") + 
  ylab("Bill Length (mm)") + 
  scale_fill_viridis_d("species") +
theme(legend.position = "none")  

```
